% Copyright 1988 by Olsen & Associates (O&A), Zurich, Switzerland.
%
%		    All Rights Reserved
%
%
% Permission to use, copy, modify, and distribute this software and its
% documentation for any purpose and without fee is hereby granted,
% provided that the above copyright notice appear in all copies, and
% that both that copyright notice and this permission notice appear in
% supporting documentation, and that all modifications of this software
% or its documentation not made by O&A or its agents are accompanied
% by a prominent notice stating who made the modifications and the date
% of the modifications.
%
% O&A DISCLAIMS ALL WARRANTIES WITH REGARD TO THIS SOFTWARE AND ITS
% DOCUMENTATION, INCLUDING ALL IMPLIED WARRANTIES OF MERCHANTABILITY AND
% FITNESS.  IN NO EVENT SHALL O&A BE LIABLE FOR ANY SPECIAL, INDIRECT OR
% CONSEQUENTIAL DAMAGES, ANY DAMAGES WHATSOEVER RESULTING FROM LOSS OF
% USE, DATA OR PROFITS, WHETHER IN AN ACTION OF CONTRACT, NEGLIGENCE OR
% OTHER TORTIOUS ACTION, ARISING OUT OF OR IN CONNECTION WITH THE USE OR
% PERFORMANCE OF THIS SOFTWARE OR ITS DOCUMENTATION.
%%%%%%%%%%%%%

Adams, C. {\em The Straight Dope.} Ballantine Books, N.Y., 1988.
\smallskip

Ad Hoc Working Group. ``Modula-2 Standard Library Definition Modules.''
{\em The MODUS Quarterly}, Issue 1, January 1985.
\smallskip

Bergin, J. and S. Greenfield. ``What Does Modula-2 Need to Fully Support 
Object Oriented Programming?'' {\em SIGPLAN Notices}, 23(3), March 1988.
\smallskip

Biagnioni, E. et al. ``OSSI -- A Portable Operating System Interface and 
Utility Library for Modula-2.'' ETH IFI Technical Bulletin No. 67. June 1986.
\smallskip

Bielak, R. ``Ada vs. Modula-2: A View From the Trenches.''
{\em ACM Sigplan Notices}, 20(12), December 1985.
\smallskip

Bilbe, C. ``In Favor of Numeric Subranges as Fundamental Types.''
{\em BSI M2WG-N169}, September 4, 1987
\smallskip

Bondy, J. ``Modula-2 Standard Library Documentation.'' {\em The MODUS
Quarterly}, Issue 1, January 1985.
\smallskip

Booch, G. ``Object-Oriented Development.'' {\em IEEE Trans. Soft. Eng.}
SE-12(2), February, 1986.
\smallskip

Borland International. {\em Proposal For Library Modules and Extensions 
to Modula-2.} Borland International. [Distributed at MODUS meeting, 
September 1985.]
\smallskip

Brooks, F.P. Jr. ``No Silver Bullet: Essence and Accidents of Software 
Engineering.'' {\em Computer}, 20(4), April 1987.
\smallskip

Bush, R. ``Library Definition Modules.'' [Private Communication]
\smallskip

Bush, R. ``Modula-2 Standard Library Rationale.'' {\em The MODUS Quarterly},
Issue 1, January 1985.
\smallskip

Cheriton, D.R.  ``The V Distributed System.'' {\em Communications of the
ACM}, 31(3), March 1988.
\smallskip

Christiansen, H.P. et al. ``More or Less Concerns About Modula-2.'' {\em
SIGPLAN Notices}, 21(9), September 1986.
\smallskip

Clark, D.C. ``The Structuring of Systems Using Upcalls.'' {\em Operating
Systems Review}, 19(5), December 1985.
\smallskip

Crawford, A.L. ``High Level Input/Output in Modula-2.'' {\em SIGPLAN
Notices}, 20(12), December 1985.
\smallskip

Czyzowic, J. and M. Iglewski. ``Implementing General Types in Modula-2.'' 
{\em ACM Sigplan
Notices}, 20(12), December 1985.
\smallskip

DeMarco, T. ``Letter re Unwarranted BSI Changes.'' {\em The MODUS
Quarterly}, Issue 8, May 1987.
\smallskip

Faulk, S.R. and D.L. Parnas.  ``On Synchronization in Hard-Real-Time Systems.''
{\em Communications of the ACM}, 31(3), March 1988.
\smallskip

Goldberg, A. and D. Robson.  {\em Smalltalk-80 The Language and its 
Implementation.} Addison-Wesley Publishing Co., 1983.
\smallskip

Granger, B.  {\em There Are No Spies.} Warner Books, N.Y., 1987.
\smallskip

Heckbert, P.S. ``Ray Tracing Jell-O Brand Gelatin.'' {\em CACM}, 31(2),
February 1988.
\smallskip

Hemmendinger, D. ``Unfair Process Scheduling in Modula-2.'' {\em SIGPLAN
Notices}, 23(3), March 1988.
\smallskip

Henry, R. ``Coroutines and Processes.'' {\em The MODUS Quarterly},
Issue 8, May 1987.
\smallskip

Henry, R. ``I/O Library Description and Rationale.'' {\em BSI IST/5/13
Document N133}, Issue 4, November 1987.
\smallskip

Hoare, C.A.R.  {\em Communicating Sequential Processes.} Prentice-Hall 
International, 1985.
\smallskip

Hoppe, J. ``Some Problems with the Specification of Standard Modules.''
{\em Modula-2 News}, Issue 0, October 1984.
\smallskip

{\em IEEE Software}. 3(6), November 1986. [Special Issue on M-2.]
\smallskip

Justice, B., S. Osborne and V. Wills. ``Building an OS with Modula-2.''
{\em The MODUS Quarterly}, Issue 7, February 1987.
\smallskip

Kepecs, J. {\em Lightweight Process User's Guide.} Sun Microsystems.
[Prerelease]
\smallskip

Kernighan, B.W. and D.M. Ritchie {\em The C Programming Language.}
Prentice-Hall, 1978.
\smallskip

Logitech.  {\em Logitech Modula-2.} Version 2.0, Logitech, Inc. 1985.
\smallskip

Logitech.  {\em Logitech Modula-2.} Version 3.0, Logitech, Inc. 1987.
\smallskip

Maugham, W.S. {\em Ashenden.} Heinemann, London, 1928.
\smallskip

Nagler, R. and J. Siegel. ``Modifications to the Standard Library Proposal.''
{\em MODUS Quarterly}, Issue 4, November 1985.
\smallskip

Nagler, R. ``Request For Comments: A Modula Standard Library Proposal.''
September 1985.  [Distributed at MODUS meeting, September 1985.]

Neumann, P.G. ``On Hierarchical Design of Computer Systems for Critical
Applications.'' {\em IEEE Trans. Soft. Eng.}, SE-12(9), September 1986.
\smallskip

Nielsen, K.W. and K. Shumate. ``Designing large real-time systems with Ada.''
{\em Communications of the ACM}, 30(8):695-715, August 1987.
\smallskip

Parnas, D.L., P.C. Clements, and D.M. Weiss. ``The Modular Structure of Complex 
Systems.'' {\em I.E.E.E.S.E.}, SE-11(3), March 1985.
\smallskip

Pase, D.M. ``System Programming in Modula-2.'' {\em ACM Sigplan Notices},
20(11), November 1985.
\smallskip

Peterson, J.L. and A. Silberschatz.  {\em Operating System Concepts.}
Addison-Wesley Publishing Company, 1983.
\smallskip

Kaplan, J.D. (Ed). {\em Dialogues of Plato.} Pocket Books, N.Y., 1959.
\smallskip
 
Polunsky, R.A.  {\em SDS-XP Extended Performance Modula-2 Software 
Development System User's Guide.} Interface Technologies, 1985.
\smallskip

{\em Proposal For Library Modules and Extensions to Modula-2.} Borland
International, 4585 Scotts Valley Drive, Scotts Valley, CA 95066. [Not
Published.]
\smallskip

Rooney, A.A. {\em Pieces of My Mind.} Avon Books, N.Y., 1984.
\smallskip

Siegel, J. ``Validation of Modula-2 Language Implementations.''
{\em The MODUS Quarterly}, Issue 1, January 1985.
\smallskip 

Stanford University.  {\em V-System 4.1 Reference Manual.} Distributed 
Systems Group, December 1983.
\smallskip

Sumner, R. ``Dynamic Module Instantiation.'' {\em The MODUS Quarterly}, 
Issue 2, April 1985.  
\smallskip

Sun Microsystems. {\em Using Modula-2 on the Sun Workstation.} Sun 
Microsystems Inc., Mountain View, CA, April 1986.
\smallskip

Swift, J. {\em Gulliver's Travels.} The John C. Winston Company, N.Y. 1930.
\smallskip

Tanenbaum, A.S.  {\em Operating Systems: Design and Implementation.}
Prentice Hall, 1987.
\smallskip

Tartan Laboratories.  {\em Modula-2 User's Guide.} Tartan Laboratories 
Inc., 1984.
\smallskip

Teitelman, W. ``A Tour Through Cedar.'' {\em IEEE Trans. Soft. Eng.}, SE-11(3), 
March 1985.
\smallskip

Twain, M. {\em Life on the Mississippi.} Bantam Books, N.Y., 1985.
\smallskip

United States Department of Defense.  {\em Reference Manual for the ADA
Programming Language.} Springer-Verlag, 1983.
\smallskip

Volition Systems. {\em Modula-2 Library Manual.} Volition Systems, Inc. 1980.
\smallskip

Ward, D. ``BSI Modula-2 Working Group Standard Concurrent Programming 
Facilities.'' {\em The MODUS Quarterly}, Issue 8, May 1987.
\smallskip

Wells, J.C. {\em Concise Esperanto And English Dictionary.} Hodder and
Stoughton Ltd., Kent, 1984.
\smallskip

Wichmann, B. and D. Hill. ``Building a Random-Number Generator.''
{\em Byte}, March 1987.
\smallskip

Xerox. {\em Mesa Language Manual.} Version 3.0 XDE3.0-3001. 
Xerox Office Systems Division, Palo Alto, CA, November 1984.
\smallskip

Xerox. {\em Mesa User's Handbook.} Version 5.0. Xerox Palo Alto 
Research Center, April 1979.
\smallskip
